Refactoring the Candidate Cancer Gene Database

### 2021-05-20 manuscript.tex
implemented comments per tim

\chapter{Introduction}
In many ways, the field of Bioinformatics exists as the result of a set of challenges rather than a discovery, as with other scientific fields. The nature of the challenges have evolved from a growth in the composition of primary data and a lacking compensatory set of technical skills to evaluate it~\cite{lawlorEngineeringBioinformaticsBuilding2015}. With the bottle neck that exists at the center of this problem, previously simple notions of collating, analyzing, and visualizing results no longer scale. As a result, a measurable area of focus is not just in the novel discovery of Biological relevance but in advancement of those engineer tools that offer new or dynamic means to solve old problems with new solutions~\cite{segalDevelopingScientificSoftware2008}. Further, the crossover of Biological and technical expertise has offered, at times, a cluttered approach to implementing these solutions, resulting in less than desirable engineering~\cite{vermaLackSoftwareEngineering2013}.

@@

In this work, we present a small but emblematic case that illustrates these shifting paradigms. With these incremental changeovers, no mater the impact, the field takes on a new character in the management of its growing data challenges. The CCGD as a concept set out to tackle some of the original problems of filtering large quantities of data. More specifically, the CCGD exists as a repository for the results of forward genetic screens in transposon mutagenesis in mice~\cite{abbottCandidateCancerGene2015a}. The nature of cancer biology offers a highly complex genetic landscape with numerous phenotypes. While many investigators start with goals to identify more focused targets in the spirit of achieving precision therapies, an essential step in this process begins with whole-sale filtering of candidates for cancer driver genes~\cite{clarkCancerGeneDiscovery2019}. Dramatic impacts have been made in the field by the identification of important cancer genes, and the addition of items like transposon screen results quality scores and strength metrics adds to the ability to make this progress. We then expand on this concept by re-implementing the same achievement within a modern framework.

@@

\begin{figure}
    \center{}
    \begin{subfigure}[t]{.3\textwidth}
        \center{}
        \includegraphics[width=\textwidth]{fig/page_header.pdf}
        \subcaption{\texttt{index.Rmd} individual page header}\label{fig:pageHeader}
    \end{subfigure}%
    \hspace{2em}
    \begin{subfigure}[t]{.3\textwidth}
        \center{}
        \includegraphics[width=\textwidth]{fig/site_header.pdf}
        \caption{\texttt{\_site.yml} file for site assembly}\label{fig:siteYaml}
    \end{subfigure}
    \hspace{2em}
    \begin{subfigure}[t]{.2\textwidth}
        \center{}
        \includegraphics[width=0.8\textwidth]{fig/yaml_illustration.pdf}
        \caption{Illustration of yaml implementation}\label{fig:yamlIllust}
    \end{subfigure}
    \caption[YAML Configuration Segments]{\textbf{YAML Configuration Segments} Pictured above are the two forms of complete YAML configuration headers utilized in Rmarkdown rendering. The central takeaway from this illustration is to recognize the heavy lifting performed by these relatively small code segments. Figure~\ref{fig:pageHeader} provides an example of the header found in every \texttt{.Rmd} file that make-up the respective pages of the site. These headers provide primarily formatting configuration for the rendering process but include broader functions as well. The header can be as simple as listing a title and Pandoc output type or include LaTex bibliography functionality and read-in style sheets like the one shown. Either way, a few lines of configuration generate wildly different outputs. Figure~\ref{fig:siteYaml} presents the YAML file which serves to assemble independent Rmarkdown files into a joined website as described in Figure~\ref{fig:renderFlow}. There is less configuration here as the only real purpose is to define the site tree for the \texttt{rmarkdown::render\_site} process.}\label{fig:yamls}
\end{figure}

\subsection{Data Integration Process}
Improvements to the data integration process were accomplished using a more simplified framework to eliminate redundancy and provide a more standardized schema. The prior implementation employed some tools that are common place, but they were implemented in a fashion that lacked simplicity at times. Also, certain proprietary widgets, such as the need for conversion in and out of the Microsoft Excel format, presented functionality and compatibility choke points that generated road blocks to maintenance~\cite{abbottCandidateCancerGene2015a}. The most substantial of the changes that addressed these issues was the transition to use of R as the data manipulation toolset. More specifically, the~\texttt{build\_table.R} script established a lean extract, transform, load (ETL) process which took the existing table data and external reference data to perform a transformation which collated old and new data. The product of this process could then be delivered to the R-based shiny app deployment generated by the~\texttt{app.R} script.

Redundancy reduction was achieved within script functionality and storage utilization. Dedicated script actions that performed house keeping functions, as seen in the~\texttt{build\_table.sh} and~\texttt{backup.sh} scripts (see Figure~\ref{fig:subroutine}), made specific effort to clear all transient data content that otherwise did not require persistent storage. This accounted for a large portion of persistent data that was previously left on the server, creating storage constraints. Additionally, script reduction was improved through a reorganization of the general subroutines within the application. A reassignment of script actions was also implemented in a more cohesive manner which allowed for both improved simplicity and modularity. This is most effectively seen in the manual control~\texttt{ccgd\_upload.sh} script as the developer has the flexibility to opt out of executing several functions within the app if run manually (see Figure~\ref{fig:subroutine}). All of this core functionality was then was automated within system scheduling using several cron jobs to sustain routine maintenance in upkeep of the application's table.

@@

The structure of a general shiny app follows a highly reduced and consistent paradigm, and ours is no different. The broad format of this work flow is \texttt{ui + server $\rightarrow$ deployment}. The server specifically has a discrete composition made up of reactive constructs of inputs, expressions, and outputs which shape the app's reactive graph. This forms, more accurately, the discrete operation of the application~\cite{wickhamMasteringShinyBook}. The assembly and movement of this graph is both emblematic and practical in displaying the modularity and improved resource utilization of the reactive elements over the previous table paradigm. Without belaboring the description of reactive programming, this model employs a declarative programming style that allows for lazy code execution. What this amounts to in application is for the developer to define loose controls at a higher level while allowing the code to passively fill these requirements. This translates to greater flexibility to both the developer and the machine as resource use can be reduced through modularity in code execution~\cite{jwlloydPracticalAdvantagesDeclarative1994}.

>>>>

\chapter{Introduction}
In many ways, the field of Bioinformatics exists as the result of a set of challenges rather than a discovery, as with other scientific fields. The nature of the challenges have evolved from a growth in the composition of primary data and a lacking compensatory set of technical skills to evaluate it~\cite{lawlorEngineeringBioinformaticsBuilding2015}. With the bottle neck that exists at the center of this problem, previously simple notions of collating, analyzing, and visualizing results no longer scale. As a result, a measurable area of focus is not just in the novel discovery of biological relevance but in advancement of those engineer tools that offer new or dynamic means to solve old problems with new solutions~\cite{segalDevelopingScientificSoftware2008}. Further, the crossover of biological and technical expertise has offered, at times, a cluttered approach to implementing these solutions, resulting in less than desirable engineering~\cite{vermaLackSoftwareEngineering2013}.

@@

In this work, we present a small but emblematic case that illustrates these shifting paradigms. With these incremental changeovers, no mater the impact, the field takes on a new character in the management of its growing data challenges. The CCGD as a concept set out to tackle some of the original problems of filtering large quantities of data. More specifically, the CCGD exists as a repository for the results of forward genetic screens in transposon mutagenesis in mice~\cite{abbottCandidateCancerGene2015a}. The nature of cancer biology offers a highly complex genetic landscape with numerous phenotypes. While many investigators start with goals to identify more focused targets in the spirit of achieving precision therapies, an essential step in this process begins with wholesale filtering of candidates for cancer driver genes~\cite{clarkCancerGeneDiscovery2019}. Dramatic impacts have been made in the field by the identification of important cancer genes, and the addition of items like transposon screen results quality scores and strength metrics adds to the ability to make this progress. We then expand on this concept by re-implementing the same achievement within a modern framework.

@@

\begin{figure}
    \center{}
    \begin{subfigure}[t]{.3\textwidth}
        \center{}
        \includegraphics[width=\textwidth]{fig/page_header.pdf}
        \subcaption{\texttt{index.Rmd} individual page header}\label{fig:pageHeader}
    \end{subfigure}%
    \hspace{2em}
    \begin{subfigure}[t]{.3\textwidth}
        \center{}
        \includegraphics[width=\textwidth]{fig/site_header.pdf}
        \caption{\texttt{\_site.yml} file for site assembly}\label{fig:siteYaml}
    \end{subfigure}
    \hspace{2em}
    \begin{subfigure}[t]{.2\textwidth}
        \center{}
        \includegraphics[width=0.8\textwidth]{fig/yaml_illustration.pdf}
        \caption{Illustration of yaml implementation}\label{fig:yamlIllust}
    \end{subfigure}
    \caption[YAML Configuration Segments]{\textbf{YAML Configuration Segments} Pictured above are the two forms of complete YAML configuration headers utilized in Rmarkdown rendering. The central takeaway from this illustration is to recognize the heavy lifting performed by these relatively small code segments. Figure~\ref{fig:pageHeader} provides an example of the header found in every \texttt{.Rmd} file that make-up the respective pages of the site. These headers provide primarily formatting configuration for the rendering process but include broader functions as well. The header can be as simple as listing a title and Pandoc output type or include LaTex bibliography functionality and read-in style sheets like the one shown. Either way, a few lines of configuration generate significantly different outputs. Figure~\ref{fig:siteYaml} presents the YAML file which serves to assemble independent Rmarkdown files into a joined website as described in Figure~\ref{fig:renderFlow}. There is less configuration here as the only real purpose is to define the site tree for the \texttt{rmarkdown::render\_site} process.}\label{fig:yamls}
\end{figure}

\subsection{Data Integration Process}
Improvements to the data integration process were accomplished using a more simplified framework to eliminate redundancy and provide a more standardized schema. The prior implementation employed some tools that are commonplace, but they were implemented in a fashion that lacked simplicity at times. Also, certain proprietary widgets, such as the need for conversion in and out of the Microsoft Excel format, presented functionality and compatibility choke points that generated road blocks to maintenance~\cite{abbottCandidateCancerGene2015a}. The most substantial of the changes that addressed these issues was the transition to use of R as the data manipulation toolset. More specifically, the~\texttt{build\_table.R} script established a lean extract, transform, load (ETL) process which took the existing table data and external reference data to perform a transformation which collated old and new data. The product of this process could then be delivered to the R-based shiny app deployment generated by the~\texttt{app.R} script.

Redundancy reduction was achieved within script functionality and storage utilization. Dedicated script actions that performed housekeeping functions, as seen in the~\texttt{build\_table.sh} and~\texttt{backup.sh} scripts (see Figure~\ref{fig:subroutine}), made specific effort to clear all transient data content that otherwise did not require persistent storage. This accounted for a large portion of persistent data that was previously left on the server, creating storage constraints. Additionally, script reduction was improved through a reorganization of the general subroutines within the application. A reassignment of script actions was also implemented in a more cohesive manner which allowed for both improved simplicity and modularity. This is most effectively seen in the manual control~\texttt{ccgd\_upload.sh} script as the developer has the flexibility to opt out of executing several functions within the app if run manually (see Figure~\ref{fig:subroutine}). All of this core functionality was then was automated within system scheduling using several cron jobs to sustain routine maintenance in upkeep of the application's table.

@@

The structure of a general shiny app follows a highly reduced and consistent paradigm, and ours is no different. The broad format of this work flow is \texttt{ui + server $\rightarrow$ deployment}. The server specifically has a discrete composition made up of reactive constructs of inputs, expressions, and outputs which shape the app's reactive graph. These elements form the discrete operation of the application~\cite{wickhamMasteringShinyBook}. The assembly and movement of this graph is both emblematic and practical in displaying the modularity and improved resource utilization of the reactive elements over the previous table paradigm. Without belaboring the description of reactive programming, this model employs a declarative programming style that allows for lazy code execution. What this amounts to in application is for the developer to define loose controls at a higher level while allowing the code to passively fill these requirements. This translates to greater flexibility to both the developer and the machine as resource use can be reduced through modularity in code execution~\cite{jwlloydPracticalAdvantagesDeclarative1994}.
